Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2021-08-30 summary.tex
updated writeup per discussion

For parameter estimation, because the popularity parameters $\{\lambda_{ik}\}$ are unidentifiable, we instead estimate the edge probabilities $p_{ij} = \lambda_{i z_j} \lambda_{j, z_i}$. The parameter estimation error is the root mean square error between the true and estimated edge probabilities, which is equivalent to the Frobenius norm of $P$:

$$\mathrm{RMSE}(\hat{P}, P) = \|\hat{P} - P\|_F.$$

@@

Given ground truth community labels, Algorithm~\ref{alg:param_est} and the MLE-based
plug-in estimators perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\) (Fig. \ref{fig:lambda_est_k}). 
We observe similar behavior when comparing errors 
in the edge probability blocks (Fig. \ref{fig:p_block_est}).

{\centering \includegraphics{summary_files/figure-latex/lambda_est_k-1}

\caption{Median and IQR RMSE for popularity vectors from Algorithm 3 (red) compared against an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
\label{fig:lambda_est_k}
\end{figure}

\begin{figure}[H]
{\centering \includegraphics{summary_files/figure-latex/lambda_est_p-1}}
\caption{Median and IQR RMSE for edge probability blocks reconstructed from the outputs of Algorithm 3 (red) compared against outputs of an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
\label{fig:p_block_est}
\end{figure}

@@

We again see community detection error trending to 0 for OSC, as well as
for SSC when \(K > 2\) (Fig.~\ref{fig:clust_err_ct_sim_imba}). The
estimation error for the parameters $\{\lambda^{k \ell}\}$ as given by
Algorithm~\ref{alg:param_est} still decays to $0$ at rate $n^{-1/2}$;  
however, the MLE-based estimators for the popularity vectors appear to
decay at a slower 
rate (Fig.~\ref{fig:lambda_est_k_imba}). 
The reduced rate of decay is removed 
if we reconstruct the edge probability blocks 
by taking the outer products of the popularity vectors 
(Fig. \ref{fig:lambda_est_p_imba}).

@@

{\centering \includegraphics{summary_files/figure-latex/lambda_est_k_imba-1}

\caption{Median and IQR RMSE from Algorithm 3 (red) compared against an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be imbalanced.}\label{fig:lambda_est_k_imba}
\end{figure}

@@

\caption{Median and IQR RMSE of edge probability blocks derived from the outputs of Algorithm 3 (red) compared against an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be imbalanced.}\label{fig:lambda_est_p_imba}
\end{figure}

@@

In the third example, we analyzed the Karantaka villages data studied by
\citet{DVN/U3BIHX_2013}. We chose the \texttt{visitgo}
networks from villages 12, 31, and 46 at the household level. The label
of interest is the religious affiliation. The networks were truncated to
religions ``1'' and ``2'', and vertices of degree 0 were removed. 
The results are displayed in Fig.~\ref{fig:households-figure} and 
Table~\ref{tab:households-table}.

>>>>

For parameter estimation, because the popularity parameters $\{\lambda_{ik}\}$ are unidentifiable, we instead estimate the edge probabilities $p_{ij} = \lambda_{i z_j} \lambda_{j, z_i}$. The parameter estimation error is the root mean square error between the true and estimated edge probabilities, which is equivalent to the Frobenius norm of $P$ divided by the number of vertices:

$$\mathrm{RMSE}(\hat{P}, P) = \frac{1}{n} \|\hat{P} - P\|_F.$$

We also note that unlike the MLE-based method \citep{307cbeb9b1be48299388437423d94bf1}, the ASE method of Algorithm 3 does not require community labels if all we are interested in is the edge probability matrix, $P$. The estimate $\hat{P}$ we obtain by combining the popularity vectors outputted by Algorithm 3 is equivalent to using the full ASE of $A$ to reconstruct the edge probability matrix.

@@

Given ground truth community labels, Algorithm~\ref{alg:param_est} and the MLE-based
plug-in estimators perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\) (Fig. \ref{fig:p_block_est}).

\begin{figure}[H]
{\centering \includegraphics{summary_files/figure-latex/lambda_est_p-1}}
\caption{Median and IQR RMSE for edge probability matrices reconstructed from the outputs of Algorithm 3 (red) compared against outputs of an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
\label{fig:p_block_est}
\end{figure}

@@

We again see community detection error trending to 0 for OSC, as well as
for SSC when \(K > 2\) (Fig.~\ref{fig:clust_err_ct_sim_imba}). 
The parameter estimation error decays at rate $n^{-1/2}$ as in the balanced communities case
(Fig. \ref{fig:lambda_est_p_imba}).

@@

@@

\caption{Median and IQR RMSE of edge probabilities derived from the outputs of Algorithm 3 (red) compared against an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be imbalanced.}
\label{fig:lambda_est_p_imba}
\end{figure}

@@

In the third example, we analyzed the Karantaka villages data studied by
\citet{DVN/U3BIHX_2013}. We chose the \texttt{visitgo}
networks from villages 12, 31, and 46 at the household level. 
In these networks, each node is a household and each edge is 
an interaction between members of pairs of households. 
The label of interest is the religious affiliation. 
The networks were truncated to religions ``1'' and ``2'', 
and vertices of degree 0 were removed. 
The villages were chosen based on there being an adequate number of nodes 
between households within each religion.

The results are displayed in Fig.~\ref{fig:households-figure} and 
Table~\ref{tab:households-table}.
